---
title: "Using the temperature-oxygen-light DVM wrapper with profile data"
output:
  html_document:
    toc: yes
    fig_width: 8
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Using the temperature-oxygen-light DVM wrapper with profile data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
library(mizer)
library(ggplot2)
library(dplyr)
library(tidyr)
library(reshape2)
```

## Overview

This vignette shows a full end-to-end example for the temperature-oxygen-light
DVM wrapper using a single site/scenario profile set. The workflow is:

1. select one profile set, for example `site = "CCE"` and `scenario = "hist"`;
2. build a `profile_data` table with depth, temperature, oxygen, light, and
   micro/mesozooplankton concentrations;
3. enable the DVM wrapper with `enable_tpo2_dvm()`;
4. run a short projection with `project_tpo2_dvm()`; and
5. plot day/night vertical biomass distributions and depth-size surface plots.

The code below will use the CSV file if it is available in the repository. If
it is not available, it falls back to a small synthetic `CCE`/`hist` example
that uses the same column schema.

## Build an example profile data set

```{r}
make_dvm_example_params <- function() {
    params <- newCommunityParams(no_w = 40, n = 0.75)
    species_params(params)$p <- 0.75
    species_params(params)$ks <- 0.2 * species_params(params)$h
    if (all(is.na(species_params(params)$k)) ||
        all(species_params(params)$k == 0)) {
        species_params(params)$k <- species_params(params)$ks
    }
    setMetabolicRate(params, p = 0.75, reset = TRUE)
}

make_synthetic_cce_hist <- function() {
    data.frame(
        site = "CCE",
        scenario = "hist",
        depth_mid_m = c(5, 15, 30, 50, 75, 110, 160, 230, 320),
        depth_top_m = c(0, 10, 20, 40, 60, 90, 130, 190, 270),
        depth_bot_m = c(10, 20, 40, 60, 90, 130, 190, 270, 370),
        T_mean = c(17, 16, 14, 12, 10, 8.5, 7.5, 6.8, 6.2),
        pO2_mean = c(18, 17, 15, 12, 9, 7.5, 6.5, 5.8, 5.2),
        zmicro_mean = c(1.2, 1.0, 0.9, 0.7, 0.5, 0.35, 0.25, 0.15, 0.1),
        zmeso_mean = c(0.08, 0.1, 0.15, 0.22, 0.3, 0.35, 0.32, 0.22, 0.12),
        I_day_rel = c(1, 0.75, 0.4, 0.15, 0.05, 0.01, 0, 0, 0),
        I_night_rel = 0
    )
}

profile_path <- "data/profiles_emblematic_sites_hist_vs_ssp585_long.csv"
if (file.exists(profile_path)) {
    profiles_all <- read.csv(profile_path, stringsAsFactors = FALSE)
    profile_data <- profiles_all |>
        filter(site == "CCE", scenario == "hist") |>
        arrange(depth_mid_m)
} else {
    profile_data <- make_synthetic_cce_hist()
}

profile_data
```

## Enable the DVM wrapper and run a projection

```{r}
params <- make_dvm_example_params()

params_dvm <- enable_tpo2_dvm(
    params,
    dvm_forcing = list(
        profile_data = profile_data,
        profile_pars = list(
            mol_to_g = 12
        )
    )
)

out_dvm <- project_tpo2_dvm(
    params_dvm,
    t_max = 30 / 365,
    dt = 1 / 365,
    effort = 0
)

profiles <- getTPO2DVMProfiles(out_dvm$params_final)
final_n <- finalN(out_dvm$sim)
head(dim(profiles$p_day))
```

## Day and night environmental and resource profiles

```{r}
site_profile_df <- profile_data |>
    select(depth_mid_m, T_mean, pO2_mean, zmicro_mean, zmeso_mean,
           I_day_rel, I_night_rel) |>
    pivot_longer(-depth_mid_m, names_to = "variable", values_to = "value")

ggplot(site_profile_df,
       aes(x = value, y = depth_mid_m, colour = variable)) +
    geom_line() +
    geom_point() +
    scale_y_reverse() +
    facet_wrap(~ variable, scales = "free_x") +
    labs(x = NULL, y = "Depth (m)",
         title = "Example CCE historical profile forcing") +
    theme_minimal() +
    theme(legend.position = "none")
```

## Helper functions for biomass profiles

```{r}
get_profile_array <- function(profiles, final_n, which_profile = c("day", "night"),
                              time_index = dim(profiles$p_day)[1]) {
    which_profile <- match.arg(which_profile)
    p_array <- if (which_profile == "day") profiles$p_day else profiles$p_night
    z_vals <- as.numeric(dimnames(p_array)[[4]])
    w_vals <- as.numeric(dimnames(final_n)[[2]])

    expand.grid(
        species = dimnames(final_n)[[1]],
        size = dimnames(final_n)[[2]],
        depth = dimnames(p_array)[[4]],
        stringsAsFactors = FALSE
    ) |>
        mutate(
            abundance = as.vector(sweep(p_array[time_index, , , , drop = FALSE][1, , , ],
                                        c(1, 2), final_n, "*")),
            weight = rep(w_vals, each = length(z_vals)),
            biomass = abundance * weight,
            depth = as.numeric(depth),
            log10_size = log10(weight),
            phase = which_profile
        )
}

choose_decade_sizes <- function(final_n) {
    w_vals <- as.numeric(dimnames(final_n)[[2]])
    decade_targets <- 10^(seq(floor(log10(min(w_vals))),
                              ceiling(log10(max(w_vals))), by = 1))
    unique(vapply(decade_targets, function(target) {
        which.min(abs(log10(w_vals) - log10(target)))
    }, integer(1)))
}

profile_day_df <- get_profile_array(profiles, final_n, "day")
profile_night_df <- get_profile_array(profiles, final_n, "night")
profile_df <- bind_rows(profile_day_df, profile_night_df)
selected_idx <- choose_decade_sizes(final_n)
selected_sizes <- as.numeric(dimnames(final_n)[[2]])[selected_idx]
```

## Day/night biomass profiles for decade size classes

```{r}
selected_profiles <- profile_df |>
    filter(weight %in% selected_sizes)

ggplot(selected_profiles,
       aes(x = biomass, y = depth, colour = phase)) +
    geom_line() +
    scale_y_reverse() +
    facet_wrap(~ signif(weight, 2), scales = "free_x") +
    labs(x = "Biomass density proxy",
         y = "Depth (m)",
         title = "Day/night biomass profiles for representative decade sizes",
         colour = "Phase") +
    theme_minimal()
```

## Depth-size surface plots

These plots use depth on the y-axis, log body size on the x-axis, and colour to
show the biomass in each size-depth bin.

```{r}
ggplot(profile_df, aes(x = log10_size, y = depth, fill = biomass)) +
    geom_raster() +
    scale_y_reverse() +
    facet_wrap(~ phase) +
    scale_fill_viridis_c(trans = "log10") +
    labs(x = expression(log[10](body~mass~(g))),
         y = "Depth (m)",
         fill = "Biomass",
         title = "Day/night biomass surfaces from the DVM wrapper") +
    theme_minimal()
```

## Realised oxygen limitation profiles

```{r}
g_real_df <- bind_rows(
    as.data.frame.table(profiles$g_real_day[dim(profiles$g_real_day)[1], , , ]) |>
        mutate(phase = "day"),
    as.data.frame.table(profiles$g_real_night[dim(profiles$g_real_night)[1], , , ]) |>
        mutate(phase = "night")
) |>
    rename(species = Var1, size = Var2, depth = Var3, g_real = Freq) |>
    mutate(depth = as.numeric(as.character(depth)),
           size = as.numeric(as.character(size)),
           log10_size = log10(size))

ggplot(g_real_df, aes(x = log10_size, y = depth, fill = g_real)) +
    geom_raster() +
    scale_y_reverse() +
    facet_wrap(~ phase) +
    labs(x = expression(log[10](body~mass~(g))),
         y = "Depth (m)",
         fill = "g",
         title = "Realised oxygen limitation by size and depth") +
    theme_minimal()
```

## Notes

- If the long-format profile CSV is available, this vignette will use the
  actual `CCE` historical profile from that file.
- If the file is absent, it falls back to a synthetic profile with the same
  columns so the workflow remains runnable.
- The surface plot is often the clearest way to inspect how the day and night
  depth distributions vary across body sizes.
