---
title: "Running the temperature-oxygen DVM wrapper from a site profile"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running the temperature-oxygen DVM wrapper from a site profile}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(mizer)
library(ggplot2)
library(dplyr)
```

## Goal

This example shows how to:

1. load one vertical forcing profile from `data/profiles_sites_hist_vs_ssp585_long.csv`,
2. enable the temperature-oxygen DVM wrapper,
3. project the integrated mizer model forward while storing day/night vertical structure,
4. plot the vertical input profiles,
5. plot the final day/night biomass-by-depth structure across size classes, and
6. plot the final integrated size spectrum seen by mizer.

## Load one site profile

```{r load-profiles}
profile_path <- if (file.exists("data/profiles_sites_hist_vs_ssp585_long.csv")) {
    "data/profiles_sites_hist_vs_ssp585_long.csv"
} else {
    system.file(
        "extdata",
        "profiles_emblematic_sites_hist_vs_ssp585_long.csv",
        package = "mizer"
    )
}

profiles <- utils::read.csv(profile_path)
site <- unique(profiles$site)[1]
scenario <- "hist"
profile_one <- profiles %>%
    filter(site == !!site, scenario == !!scenario) %>%
    arrange(depth_idx)

profile_one %>%
    select(site, scenario, depth_idx, depth_mid_m, temp_C, pO2_kPa,
           zmicro, zmeso, I_day_rel, I_night_rel) %>%
    head()
```

## Plot the vertical input profiles

```{r plot-input-profiles, fig.width=10, fig.height=7}
abiotic_df <- bind_rows(
    profile_one %>%
        transmute(depth_mid_m, variable = "Temperature (C)", phase = "shared",
                  value = temp_C),
    profile_one %>%
        transmute(depth_mid_m, variable = "Oxygen (kPa)", phase = "shared",
                  value = pO2_kPa),
    profile_one %>%
        transmute(depth_mid_m, variable = "Relative light", phase = "day",
                  value = I_day_rel),
    profile_one %>%
        transmute(depth_mid_m, variable = "Relative light", phase = "night",
                  value = I_night_rel),
    profile_one %>%
        transmute(depth_mid_m, variable = "zmicro (mol m^-3)", phase = "shared",
                  value = zmicro),
    profile_one %>%
        transmute(depth_mid_m, variable = "zmeso (mol m^-3)", phase = "shared",
                  value = zmeso)
)

ggplot(abiotic_df,
       aes(x = value, y = depth_mid_m, colour = phase, group = phase)) +
    geom_line() +
    geom_point(size = 0.8) +
    scale_y_reverse() +
    facet_wrap(~ variable, scales = "free_x", ncol = 3) +
    labs(x = NULL, y = "Depth midpoint (m)",
         title = paste("Vertical input profiles for", site, scenario)) +
    theme_bw()
```

## Enable and run the DVM wrapper

```{r run-dvm}
params <- newCommunityParams()
params_dvm <- enable_tpo2_dvm(
    params,
    profiles = profiles,
    site = site,
    scenario = scenario
)

out <- project_tpo2_dvm(
    params_dvm,
    t_max = 5,
    dt = 0.1,
    effort = 0
)

cfg <- other_params(out$params_final)$tpo2_dvm
last_step <- out$profiles_saved[[length(out$profiles_saved)]]
N_int_final <- finalN(out$sim)
```

## Reconstruct final local day/night biomass distributions

```{r reconstruct-local-biomass}
integrated_to_local <- function(N_int, p_phase, H_total, dz) {
    out <- array(0, dim = dim(p_phase), dimnames = dimnames(p_phase))
    for (k in seq_along(dz)) {
        out[, , k] <- N_int * H_total * p_phase[, , k] / dz[k]
    }
    out
}

N_day_local <- integrated_to_local(
    N_int_final,
    last_step$p_day,
    cfg$geometry$H_total,
    cfg$geometry$dz
)
N_night_local <- integrated_to_local(
    N_int_final,
    last_step$p_night,
    cfg$geometry$H_total,
    cfg$geometry$dz
)

B_day_local <- sweep(N_day_local, 2, w(params_dvm), "*")
B_night_local <- sweep(N_night_local, 2, w(params_dvm), "*")
```

## Plot the last-step day/night probability distribution by depth and size

```{r plot-local-probability, fig.width=12, fig.height=8}
make_probability_df <- function(p_local, phase_label) {
    df <- as.data.frame.table(p_local, responseName = "probability")
    names(df)[1:3] <- c("species", "size", "depth")
    df %>%
        mutate(
            size_g = as.numeric(as.character(size)),
            depth = factor(depth, levels = dimnames(p_local)[[3]]),
            depth_mid_m = cfg$geometry$z_mid[match(as.character(depth),
                                                   dimnames(p_local)[[3]])],
            phase = phase_label
        )
}

probability_df <- bind_rows(
    make_probability_df(last_step$p_day, "day"),
    make_probability_df(last_step$p_night, "night")
)

depth_grid <- seq(min(cfg$geometry$z_mid), max(cfg$geometry$z_mid), by = 1)
interpolate_probability_depth <- function(df) {
    split_df <- split(df, list(df$species, df$phase, df$size_g), drop = TRUE)
    bind_rows(lapply(split_df, function(x) {
        x <- x[order(x$depth_mid_m), ]
        data.frame(
            species = x$species[1],
            phase = x$phase[1],
            size_g = x$size_g[1],
            depth_mid_m = depth_grid,
            probability = stats::approx(
                x = x$depth_mid_m,
                y = x$probability,
                xout = depth_grid,
                rule = 2
            )$y
        )
    }))
}

probability_interp_df <- interpolate_probability_depth(probability_df)

ggplot(probability_interp_df,
       aes(x = log10(size_g), y = depth_mid_m, fill = probability)) +
    geom_raster() +
    scale_y_reverse() +
    scale_fill_viridis_c() +
    facet_grid(phase ~ species) +
    labs(x = "log10 body size (g)", y = "Interpolated depth (m)",
         fill = "Probability",
         title = "Final vertical day/night probability structure by size class") +
    theme_bw()
```

## Plot the final integrated size spectrum from mizer

```{r plot-integrated-spectrum, fig.width=10, fig.height=6}
spectrum_df <- as.data.frame.table(N_int_final, responseName = "abundance")
names(spectrum_df)[1:2] <- c("species", "size")
spectrum_df <- spectrum_df %>%
    mutate(size_g = as.numeric(as.character(size)),
           biomass_density = abundance * size_g)

ggplot(spectrum_df, aes(x = size_g, y = abundance, colour = species)) +
    geom_line() +
    scale_x_log10() +
    scale_y_log10() +
    labs(x = "Body size (g)", y = "Integrated abundance density",
         title = "Final integrated mizer size spectrum") +
    theme_bw()
```

```{r plot-integrated-biomass-spectrum, fig.width=10, fig.height=6}
ggplot(spectrum_df, aes(x = size_g, y = biomass_density, colour = species)) +
    geom_line() +
    scale_x_log10() +
    scale_y_log10() +
    labs(x = "Body size (g)", y = "Integrated biomass density",
         title = "Final integrated biomass spectrum") +
    theme_bw()
```

## Inspect the last-step realised day/night fields

```{r inspect-last-step-fields}
str(list(
    p_day = dim(last_step$p_day),
    p_night = dim(last_step$p_night),
    realized_day_nu = dim(last_step$realized_day$nu),
    realized_night_mu = dim(last_step$realized_night$mu),
    n_pp_eff = length(last_step$effective$n_pp_eff)
))
```
