---
title: "Using the temperature-oxygen community extension"
output:
  html_document:
    toc: yes
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Using the temperature-oxygen community extension}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
library(mizer)
library(ggplot2)
library(dplyr)
library(reshape2)
library(tidyr)
library(purrr)
```

## Overview

The temperature-oxygen community extension adds environmental forcing to a
standard `newCommunityParams()` model. Once enabled with
`enable_tpo2_community()`, the model keeps mizer's usual encounter dynamics,
but modifies feeding, predation and energy allocation using time-varying
temperature and dissolved oxygen forcing.

This vignette shows how to:

- enable the extension with simple forcing data;
- inspect time-varying diagnostics such as raw feeding level, realised feeding
  level and oxygen downregulation (`g`);
- plot biomass time series and size spectra with `ggplot2`; and
- explore alternative environmental scenarios with `project_tpo2()`.

## Set up a temperature-oxygen community model

We begin with a community model, but here we explicitly switch basal and active
metabolism on so the oxygen limitation term has meaningful size structure. In
particular, we set both the maximum-intake exponent `n` and the basal
metabolism exponent `p` to 0.75, give the basal metabolic coefficient `ks` a
positive value, and if `k` is zero we set the active-metabolism coefficient to
match `ks`. The forcing can be supplied either as functions of time or as
vectors with matching `times`.

```{r}
make_tpo2_community_params <- function() {
    params <- newCommunityParams(n = 0.75)

    species_params(params)$p <- 0.75
    species_params(params)$ks <- 0.2 * species_params(params)$h
    if (all(is.na(species_params(params)$k)) ||
        all(species_params(params)$k == 0)) {
        species_params(params)$k <- species_params(params)$ks
    }

    setMetabolicRate(params, p = 0.75, reset = TRUE)
}

params <- make_tpo2_community_params()

years <- seq(0, 20, by = 0.25)
forcing <- list(
    times = years,
    T_vec = 10 + 2 * sin(2 * pi * years / 5),
    pO2_vec = 18 - 4 * (years / max(years)) + 1.5 * cos(2 * pi * years / 5)
)

params_tpo2 <- enable_tpo2_community(
    params,
    forcing = forcing,
    tpo2_pars = list(
        Q10_C = 2,
        Q10_M = 2,
        Q10_A = 1.2,
        K_g = 2
    )
)
```

With this setup, basal metabolism still follows the allometric term from
`setMetabolicRate()`, active metabolism uses the linear `k * w` term scaled by
`U`, and the tpo2 extension applies the lagged oxygen multiplier `g_prev` only
to the active/SDA contribution, not to the basal component.

## A first projection

Use `project_tpo2()` when you want the lagged oxygen limitation term to be
updated across saved time steps.

```{r}
out_base <- project_tpo2(params_tpo2, t_max = 20, dt = 0.1, effort = 0)
sim_base <- out_base$sim
```

We can also inspect the forcing directly.

```{r}
env_df <- tibble(
    time = years,
    temperature = forcing$T_vec,
    oxygen = forcing$pO2_vec
) |>
    pivot_longer(cols = c(temperature, oxygen),
                 names_to = "variable", values_to = "value")

ggplot(env_df, aes(x = time, y = value, colour = variable)) +
    geom_line(linewidth = 0.8) +
    facet_wrap(vars(variable), scales = "free_y", ncol = 1) +
    labs(x = "Time [years]", y = NULL,
         title = "Environmental forcing used in the base run") +
    theme_minimal()
```

## Biomass time series

The standard mizer summaries continue to work for the `MizerSim` returned by
`project_tpo2()`.

```{r}
biomass_df <- melt(getBiomass(sim_base))

# The community model only has one species, but the same code scales to
# multi-species models.
ggplot(biomass_df, aes(x = time, y = value, colour = sp)) +
    geom_line(linewidth = 0.9) +
    labs(x = "Time [years]", y = "Biomass [g]",
         colour = "Group",
         title = "Community biomass through time") +
    theme_minimal()
```

## Size spectra

You can use the built-in `plotSpectra()` function, but it is often helpful to
build a custom `ggplot2` figure from the simulation arrays.

```{r}
plotSpectra(sim_base, power = 2)
```

```{r}
final_spectrum_df <- melt(N(sim_base)[idxFinalT(sim_base), , , drop = FALSE]) |>
    filter(value > 0)

ggplot(final_spectrum_df, aes(x = w, y = value * w^2, colour = sp)) +
    geom_line(linewidth = 0.9) +
    scale_x_log10() +
    scale_y_log10() +
    labs(x = "Size [g]", y = expression(N(w) %.% w^2),
         colour = "Group",
         title = "Final biomass spectrum under temperature-oxygen forcing") +
    theme_minimal()
```

## Diagnostic quantities: raw feeding level, realised feeding level and `g`

`getTPO2Diagnostics()` returns the main diagnostic arrays for a given model
state. To turn these into a time series, evaluate the diagnostics for each saved
time in the simulation and summarise over size.

```{r}
get_diag_summary <- function(params, sim, saved_g = NULL) {
    times <- getTimes(sim)
    n_template <- initialN(params)

    map_dfr(seq_along(times), function(i) {
        time_i <- times[i]
        params_i <- params
        if (!is.null(saved_g)) {
            other <- other_params(params_i)
            other$tpo2$g_prev <- saved_g[i, , ]
            other_params(params_i) <- other
        }

        n_i <- N(sim)[i, , ]
        if (is.null(dim(n_i))) {
            n_i <- array(n_i,
                         dim = dim(n_template),
                         dimnames = dimnames(n_template))
        }

        diag_i <- getTPO2Diagnostics(
            params_i,
            n = n_i,
            n_pp = NResource(sim)[i, ],
            n_other = NOther(sim)[i, ],
            t = time_i
        )

        tibble(
            time = time_i,
            raw_feeding_level = weighted.mean(diag_i$f, w = as.numeric(n_i)),
            realised_feeding_level = weighted.mean(diag_i$f_real,
                                                   w = as.numeric(n_i)),
            g = weighted.mean(diag_i$g, w = as.numeric(n_i)),
            oxygen_intake = weighted.mean(diag_i$C,
                                          w = as.numeric(n_i)),
            active_metabolism = weighted.mean(pmax(0, diag_i$M_A),
                                              w = as.numeric(n_i))
        )
    })
}

diag_base_df <- get_diag_summary(params_tpo2, sim_base, out_base$saved_g) |>
    pivot_longer(cols = -time, names_to = "metric", values_to = "value")

ggplot(diag_base_df, aes(x = time, y = value, colour = metric)) +
    geom_line(linewidth = 0.9) +
    facet_wrap(vars(metric), scales = "free_y", ncol = 1) +
    labs(x = "Time [years]", y = NULL,
         title = "Time-varying temperature-oxygen diagnostics") +
    theme_minimal() +
    theme(legend.position = "none")
```

The distinction between `f` and `f_real` is important:

- `f` is the raw feeding level implied by encounter rate and temperature-adjusted
  intake;
- `g` is the oxygen downregulation scalar; and
- `f_real = f * g` is the realised feeding level after oxygen limitation.

A size-resolved diagnostic view is often useful as well.

```{r}
diag_final <- getTPO2Diagnostics(
    out_base$params_final,
    t = max(getTimes(sim_base))
)

diag_size_df <- bind_rows(
    melt(diag_final$f) |> mutate(metric = "raw feeding level"),
    melt(diag_final$f_real) |> mutate(metric = "realised feeding level"),
    melt(diag_final$g) |> mutate(metric = "oxygen downregulation g")
) |>
    rename(size = w, species = sp)

ggplot(diag_size_df, aes(x = size, y = value, colour = metric)) +
    geom_line(linewidth = 0.9) +
    scale_x_log10() +
    labs(x = "Size [g]", y = NULL,
         title = "Size-dependent diagnostics at the final time") +
    theme_minimal()
```

## Size-specific growth and reproduction context

Because the extension changes energy available for growth and reproduction, it
is useful to compare the usual mizer summaries alongside the new diagnostics.

```{r}
growth_df <- melt(getEGrowth(out_base$params_final)) |>
    rename(size = w, species = sp)

repro_df <- melt(getERepro(out_base$params_final)) |>
    rename(size = w, species = sp)

ggplot(growth_df, aes(x = size, y = value, colour = species)) +
    geom_line(linewidth = 0.9) +
    scale_x_log10() +
    labs(x = "Size [g]", y = "Energy for growth",
         title = "Growth energy across the size spectrum") +
    theme_minimal()
```

```{r}
ggplot(repro_df, aes(x = size, y = value, colour = species)) +
    geom_line(linewidth = 0.9) +
    scale_x_log10() +
    labs(x = "Size [g]", y = "Energy for reproduction",
         title = "Reproductive allocation across the size spectrum") +
    theme_minimal()
```

## Scenario exploration

A convenient workflow is to define a small helper that constructs forcing,
enables the extension and runs `project_tpo2()`. That makes it easy to compare
multiple climate or oxygen scenarios.

```{r}
run_tpo2_scenario <- function(name, temperature, oxygen, effort = 0) {
    forcing_i <- list(
        times = years,
        T_vec = temperature,
        pO2_vec = oxygen
    )

    params_i <- enable_tpo2_community(make_tpo2_community_params(), forcing = forcing_i)
    out_i <- project_tpo2(params_i, t_max = max(years), dt = 0.1, effort = effort)

    list(
        name = name,
        params = params_i,
        out = out_i
    )
}

scenarios <- list(
    run_tpo2_scenario(
        "Baseline",
        temperature = 10 + 2 * sin(2 * pi * years / 5),
        oxygen = 18 - 4 * (years / max(years)) + 1.5 * cos(2 * pi * years / 5)
    ),
    run_tpo2_scenario(
        "Warm, low oxygen",
        temperature = 12 + 2.5 * sin(2 * pi * years / 5),
        oxygen = 15 - 5 * (years / max(years))
    ),
    run_tpo2_scenario(
        "Cool, oxygen-rich",
        temperature = 9 + 1.5 * sin(2 * pi * years / 5),
        oxygen = 19 + 1 * cos(2 * pi * years / 5)
    )
)
```

### Compare biomass trajectories

```{r}
scenario_biomass_df <- map_dfr(scenarios, function(x) {
    melt(getBiomass(x$out$sim)) |>
        mutate(scenario = x$name)
})

ggplot(scenario_biomass_df,
       aes(x = time, y = value, colour = scenario, linetype = scenario)) +
    geom_line(linewidth = 0.9) +
    labs(x = "Time [years]", y = "Biomass [g]",
         title = "Biomass response under alternative scenarios") +
    theme_minimal()
```

### Compare size spectra at the final time

```{r}
scenario_spectra_df <- map_dfr(scenarios, function(x) {
    melt(N(x$out$sim)[idxFinalT(x$out$sim), , , drop = FALSE]) |>
        filter(value > 0) |>
        mutate(scenario = x$name)
})

ggplot(scenario_spectra_df,
       aes(x = w, y = value * w^2, colour = scenario, linetype = scenario)) +
    geom_line(linewidth = 0.9) +
    scale_x_log10() +
    scale_y_log10() +
    labs(x = "Size [g]", y = expression(N(w) %.% w^2),
         title = "Final size spectra across scenarios") +
    theme_minimal()
```

### Compare realised feeding level and `g`

```{r}
scenario_diag_df <- map_dfr(scenarios, function(x) {
    get_diag_summary(x$params, x$out$sim, x$out$saved_g) |>
        mutate(scenario = x$name)
}) |>
    pivot_longer(cols = c(raw_feeding_level, realised_feeding_level, g),
                 names_to = "metric", values_to = "value")

ggplot(scenario_diag_df,
       aes(x = time, y = value, colour = scenario, linetype = scenario)) +
    geom_line(linewidth = 0.9) +
    facet_wrap(vars(metric), scales = "free_y", ncol = 1) +
    labs(x = "Time [years]", y = NULL,
         title = "Scenario comparison for feeding and oxygen limitation") +
    theme_minimal()
```

These plots make it straightforward to ask practical scenario questions, for
example:

- How much does declining oxygen reduce realised feeding level?
- Do warmer conditions increase metabolic pressure enough to suppress biomass?
- Which sizes are most affected when `g` falls well below 1?

## Where to go next

This community example keeps the model intentionally simple. The same workflow
extends naturally to larger mizer models:

1. start from a calibrated `MizerParams` object;
2. enable the temperature-oxygen extension with observed or scenario forcing;
3. run `project_tpo2()`;
4. summarise `getTPO2Diagnostics()` over time, species or size; and
5. compare scenarios with the usual mizer summaries such as biomass, spectra,
   feeding level, growth and reproduction.

That combination of standard mizer outputs and explicit temperature-oxygen
diagnostics is usually enough to build a transparent scenario-analysis workflow.
